---
title: "Final Project"
author: "Brittany O-W and Molly F-F"
date: "2024-12-04"
output: html_document
---

Link to GitHub repository: https://github.com/bnowen/FischFriedman-Owen-FinalProject

#Final Post
We explore how Google search volume related to abortion, birth control, and reproductive healthcare changed before and after the 2024 presidential election in the US by state and metropolitan area. We use the gtrendsR package to extract search history data for related terms through the Google Trends API. Our unit of analysis is search result density and any changes across our tracked time periods. We track search results by metro area (N=210) and state (N=51) for an additional, larger-scale perspective, as several states had abortion measures on the ballot for this election. Analysis was conducted on the Top 5 terms with the largest relative search volume after considering roughly 20 integer variables. Our initial, exploratory analysis was conducted using plots from base R and the ggplot2 package. From there, we utilized ggplot2 and plotly to create two interactive bubble plot charts that allow users to compare a search term (the color of a bubble) by its search volume (size of the bubble and x-axis) for each region or state (y-axis) over time in two-day increments (secondary x-axis). Users are able to hover over each bubble for more information on the exact search volume, location, and date. 

Figures can be downloaded and viewed as local .html files from this repository:
https://github.com/bnowen/FischFriedman-Owen-SURV727-Figures

Note: If not all states or metropolitan areas are visible on the y-axis due to image resolution, please zoom in so that the figure can populate the entire space. To view static data for one date, click on the desired date in the slider and the figure will remain still.

#Data Pull

```{r}
library(tidyverse)
library(gtrendsR)
library(censusapi)
library(dplyr)
```

```{r}
#Checking for popular topics

searches1 <- gtrends(c("abortion", "birth control", "IUD", "tubes tied", "hysterectomy"),
               geo = "US",
               time = "2024-10-23 2024-11-20",
               low_search_volume = TRUE)
plot(searches1)

searches1_time <- as_tibble(searches1$interest_over_time)
searches1_region <- as_tibble(searches1$interest_by_region)
searches1_metro <- as_tibble(searches1$interest_by_dma)

saveRDS(searches1_time, "searches1_time.rds")
saveRDS(searches1_region, "searches1_region.rds")
saveRDS(searches1_metro, "searches1_metro.rds")


```
```{r}
searches2 <- gtrends(c("vasectomy", "family planning", "plan B", "birth control implant"),
               geo = "US",
               time = "2024-10-23 2024-11-20",
               low_search_volume = TRUE)
plot(searches2)

searches2_time <- as_tibble(searches2$interest_over_time)
searches2_region <- as_tibble(searches2$interest_by_region)
searches2_metro <- as_tibble(searches2$interest_by_dma)

saveRDS(searches2_time, "searches2_time.rds")
saveRDS(searches2_region, "searches2_region.rds")
saveRDS(searches2_metro, "searches2_metro.rds")
```
```{r}
searches3 <- gtrends(c("tubal litigation", "bisalp", "abortion clinic", "abortion center"),
               geo = "US",
               time = "2024-10-23 2024-11-20",
               low_search_volume = TRUE)
plot(searches3)

searches3_time <- as_tibble(searches3$interest_over_time)
searches3_region <- as_tibble(searches3$interest_by_region)
searches3_metro <- as_tibble(searches3$interest_by_dma)

saveRDS(searches3_time, "searches3_time.rds")
saveRDS(searches3_region, "searches3_region.rds")
saveRDS(searches3_metro, "searches3_metro.rds")
```

# Looking at most common terms

``` {r}
searches1_time$hits <- as.numeric(searches1_time$hits)
ggplot(searches1_time, aes(x=date, y=hits, group=keyword)) + geom_line(aes(color=keyword))

searches2_time$hits <- as.numeric(searches2_time$hits)
ggplot(searches2_time, aes(x=date, y=hits, group=keyword)) + geom_line(aes(color=keyword))

searches3_time$hits <- as.numeric(searches3_time$hits)
ggplot(searches3_time, aes(x=date, y=hits, group=keyword)) + geom_line(aes(color=keyword))
```

Abortion and Plan B were the most common search terms in their respective groups, each peaking on November 6, the day after the election. Abortion clinic was the most common search term in its group, but it peaked before the election.

#Reshape Data

```{r}
searches1_metro_w <- spread(searches1_metro, key = keyword, value = hits)
searches1_region_w <- spread(searches1_region, key = keyword, value = hits)
searches1_time_w <- spread(searches1_time, key = keyword, value = hits)

searches2_metro_w <- spread(searches2_metro, key = keyword, value = hits)
searches2_region_w <- spread(searches2_region, key = keyword, value = hits)
searches2_time_w <- spread(searches2_time, key = keyword, value = hits)

searches3_metro_w <- spread(searches3_metro, key = keyword, value = hits)
searches3_region_w <- spread(searches3_region, key = keyword, value = hits)
searches3_time_w <- spread(searches3_time, key = keyword, value = hits)
```

#Create final datasets for overall seaches

```{r}

#Metro area dataset
final_metro <- searches1_metro_w
library(dplyr)
final_metro <- left_join(final_metro, searches2_metro_w[,c("location", "birth control implant","family planning", "vasectomy")], by="location")
final_metro <- left_join(final_metro, searches3_metro_w[,c("location", "abortion center","abortion clinic", "bisalp", "tubal litigation")], by="location")
```
```{r}
#region area dataset
final_region <- searches1_region_w
final_region <- left_join(final_region, searches2_region_w[,c("location", "birth control implant","family planning", "vasectomy")], by="location")
final_region <- left_join(final_region, searches3_region_w[,c("location", "bisalp", "tubal litigation")], by="location")

```
```{r}
#time area dataset
final_time <- searches1_time_w
final_time <- left_join(final_time, searches2_time_w[,c("date", "birth control implant","family planning", "vasectomy")], by="date")
final_time <- left_join(final_time, searches3_time_w[,c("date", "bisalp", "tubal litigation")], by="date")

```
```{r}
saveRDS(final_metro, "final_metro.rds")
saveRDS(final_region, "final_region.rds")
saveRDS(final_time, "final_time.rds")
```

#Time Series - This takes a while to run, be cautious 
```{r}
oct23 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
               geo = "US",
               time = "2024-10-23 2024-10-24",
               low_search_volume = TRUE)
#oct24 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              # geo = "US",
               #time = "2024-10-24 2024-10-25",
               #low_search_volume = TRUE)
oct25 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
               geo = "US",
               time = "2024-10-25 2024-10-26",
               low_search_volume = TRUE)
#oct26 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              # geo = "US",
               #time = "2024-10-26 2024-10-27",
               #low_search_volume = TRUE)
oct27 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
               geo = "US",
               time = "2024-10-27 2024-10-28",
               low_search_volume = TRUE)
#oct28 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
            #  geo = "US",
             #  time = "2024-10-28 2024-10-29",
              # low_search_volume = TRUE)
oct29 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-10-29 2024-10-30",
               low_search_volume = TRUE)
#oct30 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-10-30 2024-10-31",
   #            low_search_volume = TRUE)
oct31 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-10-31 2024-11-01",
               low_search_volume = TRUE)
#nov1 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-01",
   #            low_search_volume = TRUE)
nov2 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-02 2024-11-03",
               low_search_volume = TRUE)
#nov3 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-03",
   #            low_search_volume = TRUE)
nov4 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-04 2024-11-05",
               low_search_volume = TRUE)
#nov5 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-05",
   #            low_search_volume = TRUE)
nov6 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-06 2024-11-07",
               low_search_volume = TRUE)
#nov7 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-07",
   #            low_search_volume = TRUE)
nov8 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-08 2024-11-09",
               low_search_volume = TRUE)
#nov9 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-09",
   #            low_search_volume = TRUE)
nov10 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-10 2024-11-11",
               low_search_volume = TRUE)
#nov11 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-11",
   #            low_search_volume = TRUE)
nov12 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-12 2024-11-13",
               low_search_volume = TRUE)
#nov13 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-13",
   #            low_search_volume = TRUE)
nov14 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-14 2024-11-15",
               low_search_volume = TRUE)
#nov15 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-15",
   #            low_search_volume = TRUE)
nov16 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-16 2024-11-17",
               low_search_volume = TRUE)
#nov17 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-17",
   #            low_search_volume = TRUE)
nov18 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
              geo = "US",
               time = "2024-11-18 2024-11-19",
               low_search_volume = TRUE)
#nov19 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-19",
   #            low_search_volume = TRUE)
#nov20 <- gtrends(c("abortion", "bisalp", "plan B", "abortion clinic", "vasectomy"),
 #             geo = "US",
  #             time = "2024-11-20",
   #            low_search_volume = TRUE)



```

```{r}

oct23_metro <- as_tibble(oct23$interest_by_dma)
oct25_metro <- as_tibble(oct25$interest_by_dma)
oct27_metro <- as_tibble(oct27$interest_by_dma)
oct29_metro <- as_tibble(oct29$interest_by_dma)
oct31_metro <- as_tibble(oct31$interest_by_dma)
nov2_metro <- as_tibble(nov2$interest_by_dma)
nov4_metro <- as_tibble(nov4$interest_by_dma)
nov6_metro <- as_tibble(nov6$interest_by_dma)
nov8_metro <- as_tibble(nov8$interest_by_dma)
nov10_metro <- as_tibble(nov10$interest_by_dma)
nov12_metro <- as_tibble(nov12$interest_by_dma)
nov14_metro <- as_tibble(nov14$interest_by_dma)
nov16_metro <- as_tibble(nov16$interest_by_dma)
nov18_metro <- as_tibble(nov18$interest_by_dma)

oct23_region <- as_tibble(oct23$interest_by_region)
oct25_region <- as_tibble(oct25$interest_by_region)
oct27_region <- as_tibble(oct27$interest_by_region)
oct29_region <- as_tibble(oct29$interest_by_region)
oct31_region <- as_tibble(oct31$interest_by_region)
nov2_region <- as_tibble(nov2$interest_by_region)
nov4_region <- as_tibble(nov4$interest_by_region)
nov6_region <- as_tibble(nov6$interest_by_region)
nov8_region <- as_tibble(nov8$interest_by_region)
nov10_region <- as_tibble(nov10$interest_by_region)
nov12_region <- as_tibble(nov12$interest_by_region)
nov14_region <- as_tibble(nov14$interest_by_region)
nov16_region <- as_tibble(nov16$interest_by_region)
nov18_region <- as_tibble(nov18$interest_by_region)


```

```{r}
oct23_region$date = "2024-10-23 2024-10-24"
oct25_region$date = "2024-10-25 2024-10-26"
oct27_region$date = "2024-10-27 2024-10-28"
oct29_region$date = "2024-10-29 2024-10-30"
oct31_region$date = "2024-10-31 2024-11-01"
nov2_region$date = "2024-11-02 2024-11-03"
nov4_region$date = "2024-11-04 2024-11-05"
nov6_region$date = "2024-11-06 2024-11-07"
nov8_region$date = "2024-11-08 2024-11-09"
nov10_region$date = "2024-11-10 2024-11-11"
nov12_region$date = "2024-11-12 2024-11-13"
nov14_region$date = "2024-11-14 2024-11-15"
nov16_region$date = "2024-11-16 2024-11-17"
nov18_region$date = "2024-11-18 2024-11-19"
```

```{r}
byregion <- rbind(oct23_region, oct25_region, oct27_region, oct29_region, oct31_region, nov2_region, nov4_region, nov6_region, nov8_region, nov10_region, nov12_region, nov14_region, nov16_region, nov18_region)
saveRDS(byregion, "byregion.rds")
```

```{r}
oct23_metro$date = "2024-10-23 2024-10-24"
oct25_metro$date = "2024-10-25 2024-10-26"
oct27_metro$date = "2024-10-27 2024-10-28"
oct29_metro$date = "2024-10-29 2024-10-30"
oct31_metro$date = "2024-10-31 2024-11-01"
nov2_metro$date = "2024-11-02 2024-11-03"
nov4_metro$date = "2024-11-04 2024-11-05"
nov6_metro$date = "2024-11-06 2024-11-07"
nov8_metro$date = "2024-11-08 2024-11-09"
nov10_metro$date = "2024-11-10 2024-11-11"
nov12_metro$date = "2024-11-12 2024-11-13"
nov14_metro$date = "2024-11-14 2024-11-15"
nov16_metro$date = "2024-11-16 2024-11-17"
nov18_metro$date = "2024-11-18 2024-11-19"
```

```{r}
bymetro <- rbind(oct23_metro, oct25_metro, oct27_metro, oct29_metro, oct31_metro, nov2_metro, nov4_metro, nov6_metro, nov8_metro, nov10_metro, nov12_metro, nov14_metro, nov16_metro, nov18_metro)
saveRDS(bymetro, "bymetro.rds")
```

```{r}
write.csv(byregion, "byregion.csv", row.names=FALSE)

```


#Interactive Plot

```{r}
library(ggplot2)
library(psych)
library(tidyr)
library(plotly)
library(extrafont)
library(htmlwidgets)
```

#FOR MOLLY
#Plotly
https://plotly.com/r/animations/
https://plotly.com/r/creating-and-updating-figures/

NOTES:

- But it seems to have a problem with "bisalp". From what I've read online, this seems to be because "bisalp" isn't present on all search dates (earlier before the election). --> FIXED: set na to -1

- Date variable needs to be set to DATETIME format or Plotly cannot format it correctly --> fixed

- can we figure out a way to make the plot longer so that every state is written on the y axis?

```{r}
byregion[is.na(byregion)] <- -1

forplot <- byregion

forplot <- forplot %>% 
     mutate(date = recode(date, '2024-10-23 2024-10-24' = as.Date('2024-10-23'), '2024-10-25 2024-10-26' = as.Date('2024-10-25'), '2024-10-27 2024-10-28' = as.Date('2024-10-27'), '2024-10-29 2024-10-30' = as.Date('2024-10-29'), '2024-10-31 2024-11-01' = as.Date('2024-10-31'), '2024-11-02 2024-11-03' = as.Date('2024-11-02'), '2024-11-04 2024-11-05' = as.Date('2024-11-04'), '2024-11-06 2024-11-07' = as.Date('2024-11-06'), '2024-11-08 2024-11-09' = as.Date('2024-11-08'), '2024-11-10 2024-11-11' = as.Date('2024-11-10'), '2024-11-12 2024-11-13' = as.Date('2024-11-12'), '2024-11-14 2024-11-15' = as.Date('2024-11-14'), '2024-11-16 2024-11-17' = as.Date('2024-11-16'), '2024-11-18 2024-11-19' = as.Date('2024-11-18')))
  

```
```{r}
library(plotly)
fig <- plot_ly(forplot, x = ~hits, y = ~location, size = ~hits, color = ~keyword, frame= ~date, type = 'scatter', mode = "markers", fill = ~'')

fig <- fig %>% layout(
    title = "Search Results by Top 5 Keywords",
    yaxis = list(title = "State"),
  legend=list(title=list(text='Keyword')),
  xaxis = list(
    title = "Hits",
    tickmode = "array"
  ),
    margin = list(l = 100)
  )

fig <- fig %>% animation_slider(
    currentvalue = list(prefix = "Date: ", font = list(color="black"))
  )

fig <- fig %>%
  animation_opts( #higher value makes animation slower
    2000, easing = "cubic-in", redraw = FALSE 
  )
fig
```


#Save plot as .png - can also save using plotly taskbar in plot
```{r}
fig <- ggsave("interactive_plot.png")
```

#Save as html gif
```{r}
htmlwidgets::saveWidget(as_widget(fig), "index.html")
```


# same as above but for metro area

```{r}
bymetro[is.na(bymetro)] <- -1

forplot2 <- bymetro

forplot2 <- forplot2 %>% 
     mutate(date = recode(date, '2024-10-23 2024-10-24' = as.Date('2024-10-23'), '2024-10-25 2024-10-26' = as.Date('2024-10-25'), '2024-10-27 2024-10-28' = as.Date('2024-10-27'), '2024-10-29 2024-10-30' = as.Date('2024-10-29'), '2024-10-31 2024-11-01' = as.Date('2024-10-31'), '2024-11-02 2024-11-03' = as.Date('2024-11-02'), '2024-11-04 2024-11-05' = as.Date('2024-11-04'), '2024-11-06 2024-11-07' = as.Date('2024-11-06'), '2024-11-08 2024-11-09' = as.Date('2024-11-08'), '2024-11-10 2024-11-11' = as.Date('2024-11-10'), '2024-11-12 2024-11-13' = as.Date('2024-11-12'), '2024-11-14 2024-11-15' = as.Date('2024-11-14'), '2024-11-16 2024-11-17' = as.Date('2024-11-16'), '2024-11-18 2024-11-19' = as.Date('2024-11-18')))
  

```
```{r}
library(plotly)
fig2 <- plot_ly(forplot2, x = ~hits, y = ~location, size = ~hits, color = ~keyword, frame= ~date, type = 'scatter', mode = "markers", fill = ~'')

fig2 <- fig2 %>% layout(
    title = "Search Results by Top 5 Keywords",
    yaxis = list(title = "Metro Area"),
  legend=list(title=list(text='Keyword')),
  xaxis = list(
    title = "Hits",
    tickmode = "array"
  ),
    margin = list(l = 100)
  )

fig2 <- fig2 %>% animation_slider(
    currentvalue = list(prefix = "Date: ", font = list(color="black"))
  )

fig2 <- fig2 %>%
  animation_opts( #higher value makes animation slower
    2000, easing = "cubic-in", redraw = FALSE 
  )
fig2
```


#Save plot as .png - can also save using plotly taskbar in plot
```{r}
fig2 <- ggsave("interactive_plot2.png")
```

#Save as html gif
```{r}
htmlwidgets::saveWidget(as_widget(fig2), "index.html")
```
